---
title: "cs07: One Script, Many Products"
format: 
  html: default
  gfm: default
  pptx: default
  docx: default
---

## Mauna Loa CO~2~ Data
The data for this case study is sourced from the NOAA Earth System Research Laboratory’s Global Monitoring Laboratory. Mauna Loa Annual Mean CO~2~ data is available for download [here](ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_annmean_mlo.txt). 

```{r load libraries and data}
#| include: false
#| warning: false
library(quarto)
library(tidyverse)
library(ggtext)
library(knitr)

data_url <- "ftp://aftp.cmdl.noaa.gov/products/trends/co2/co2_annmean_mlo.txt"

co2 <- readr::read_table(data_url, comment = "#", col_names = c("year", "mean", "unc"))

```

### Time Series of CO~2~ Concentrations
The following graph shows annual mean CO~2~ concentrations measured at Mauna Loa Observatory from 1959 through 2024. 


``` {r time series plot}
#| echo: false
co2_time <- ggplot(data = co2, aes(x = year, y = mean)) +
  geom_line(color = "darkslateblue", linewidth = 1.5) +
  theme_bw() +
  theme(axis.title.y = element_markdown(),
        plot.title = element_markdown()) +
  labs(x = "Year",
       y = "CO<sub>2</sub> (ppm)",
       title = "Annual Mean CO<sub>2</sub> Concentrations at Mauna Loa")
co2_time
```

### Mean CO~2~ by Decade
The following table shows mean CO~2~ concentrations by decade. Note that the decade beginning in 2020 only includes observations through 2024. 

``` {r table summary}
#| echo: false

co2_decade <- co2 |>
  slice(-1) |>
  mutate(decade = floor(year / 10) * 10) |>
  group_by(decade) |>
  summarize(mean(mean))

kable(co2_decade, format = "html", col.names = c("Decade Start Year", "Mean CO<sub>2</sub> (ppm)"),
      align = "c", digits = round(1), caption = "Mean CO<sub>2</sub> Concentrations by Decade",
      escape = FALSE)
```

